---
title: "Passion Driven Statistics"
author: "Alan T. Arnholt"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Passion Driven Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
options(useFancyQuotes="UTF-8")
knitr::opts_chunk$set(comment = NA, message = FALSE, warning = FALSE, prompt = TRUE)
```


## Using **PDS**

Once the `PDS` package is installed, you will need to load the package using the `library` function.   

```{r}
library(PDS)
```

Now it is possible to access the data in the four data sets by typing the name of the data frame at the `R` prompt.  For example, to see the first five values in the variable `incomeperperson` from the `gapminder` data frame enter

```{r}
gapminder[1:5, 'incomeperperson']
```

### Data Management Using the `dplyr` Package

The package `dplyr` must first be installed before it can be used.  There are two ways one can install the package.  The easiest way to install `dplyr`, is to type 
```{r eval = FALSE}
install.packages("dplyr")
```
at the `R` prompt.  If you want the latest version of `dplyr`, you can install the development version from GitHub provided you have the required tools to build `R` packages installed on your machine (`Rtools` for Windows users, `XCode` for Mac users, etc.).  

```{r eval = FALSE}
if (packageVersion("devtools") < 1.6) {
  install.packages("devtools")
}
devtools::install_github("hadley/lazyeval")
devtools::install_github("hadley/dplyr")
```

Create a subset of people 25 or younger who smoked in the last 12 months.  To do this, one can use the `dplyr` package. Note that `CHECK321 == 1` is used to see if subject has smoked in the past 12 months and `!is.na()` is used to make sure the subset does not contain `NA`s for the chosen variables.

```{r}
library(dplyr)
nesarc <- NESARC %>% 
  filter(!is.na(CHECK321) & !is.na(AGE) & CHECK321 ==1 & AGE <= 25)
dim(nesarc)
```

The `nesarc` data frame contains `r dim(nesarc)[2]` variables.  Use the verb `select` from the `dplyr` package to select the variables `CHECK321`, `AGE` (age of subject), `TAB12MDX` (tobaco dependence past 12 months), `S3AQ3B1` (usual smoking quantity), `ETHRACE2A` (ethnicity), `SEX` (gender), `MAJORDEPLIFE` (Dysthymia-Lifetime) and `S3AQ3C1` (usual smoking frequency).  

### Renaming Variables

Renaming the variables `IDNUM`, `ETOTLCA2`, and `ETHRACE2A` to `UniqueID`, `EthanolConsumption`, and `Ethnicity`, respectively, is done with the `rename` function.  Although it is not a requirement, the data frame `NESARC` is converted to a `data frame tbl` per the advice given in the `dplyr` vignette.  The `rename` function takes a data frame or data frame tbl as its first argument while the syntax `new_name = old_name` is used as the second argument.

```{r label = "RENAME"}
library(dplyr)
NESARC[1:5, 1:5]  # Show first 5 rows and first 5 columns of NESARC
NESARCtbl <- tbl_df(NESARC) %>% 
  rename(UniqueID = IDNUM, EthanolConsumption = ETOTLCA2, Ethnicity = ETHRACE2A, 
         SmokingFrequency = S3AQ3B1, Age = AGE, MajorDepression = MAJORDEPLIFE, 
         DysthymiaLifetime = DYSLIFE, TobaccoDependence = TAB12MDX, DailyCigsSmoked = S3AQ3C1) %>% 
  select(UniqueID, EthanolConsumption, Ethnicity, SmokingFrequency, Age, MajorDepression, 
         DysthymiaLifetime, TobaccoDependence, DailyCigsSmoked)  
NESARCtbl
```

#### kable

```{r, echo = FALSE, results = 'asis'}
knitr::kable(NESARCtbl[1300:1305, c("EthanolConsumption", "SmokingFrequency", "MajorDepression")], 
             align = c("c","c","c"), caption = "Three Selected Columns")
```

#### pander

```{r, echo=FALSE, results='asis'}
pander::pandoc.table(NESARCtbl[1300:1305, c("EthanolConsumption", "SmokingFrequency", "MajorDepression")],
                     caption = "Three Selected Columns")
```

### Coding Missing Values

Note that the variable `S3AQ3B1` renamed to `SmokingFrequency` uses a 9 to record `unknown` for smoking frequency.

```{r label = "CodeMissing"}
NESARCtbl$SmokingFrequency[NESARCtbl$SmokingFrequency == 9] <- NA
summary(NESARCtbl$SmokingFrequency)  # Note that 9 still appears
NESARCtbl$SmokingFrequency <- factor(NESARCtbl$SmokingFrequency)[, drop = TRUE]
summary(NESARCtbl$SmokingFrequency)  # Unused level no longer appears
```

I hate to look back at the code book to figure out what the heck the numbers represent in this case!  
One of the great things about the `R` language is how it handles a wide variety of data types.  
In this case, the variable `SmokingFrequency` is a factor with numeric labels for the levels of the factor.  
To change the numeric labels to text type the following:

```{r label = "Freq"}
NESARCtbl$SmokingFrequency <- factor(NESARCtbl$SmokingFrequency, 
                                     labels = c("Every Day", "5 to 6 Days/week", 
                                                "3 to 4 Days/week", "1 to 2 Days/week", 
                                                "2 to 3 Days/month", "Once a month or less"))
summary(NESARCtbl$SmokingFrequency)
xtabs(~SmokingFrequency, data = NESARCtbl) # Note how the NA's are not printed
```

The following `ggplot2` code is used to graph the values in `SmokingFrequency`.

```{r label = "ggSmoke"}
library(ggplot2)
ggplot(data = NESARCtbl, aes(x = SmokingFrequency)) + 
  geom_bar(fill = "lightgray") + 
  labs(x = "Smoking Frequency") +
  theme_bw() 
```

If you do not want to see the `NA`'s or have the labels on the $x-$ axis overlap, use

```{r label = "ggSmoke2", fig.width = 6, fig.height = 6}
ggplot(data = na.omit(NESARCtbl[ , "SmokingFrequency", drop = FALSE]), aes(x = SmokingFrequency)) + 
  geom_bar(fill = "lightgray") + 
  labs(x = "Smoking Frequency") +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 85, vjust = 0.5)) 
```



Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` setion of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10, main ="Basic Plot", pch = 19, col ="red")
plot(10:1, xlab = "", ylab ="", col = "green", pch = 19)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

```{r, fig.show = 'hold', fig.cap = "Basic plot in case you wondered"}
plot(1:10, main ="Basic Plot", pch = 19, col ="red")
```

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

```{r}
1:10
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
