---
title: "Passion Driven Statistics"
author: "Alan T. Arnholt"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Passion Driven Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
options(useFancyQuotes="UTF-8")
knitr::opts_chunk$set(comment = NA, message = FALSE, warning = FALSE, prompt = TRUE)
```


## Using **PDS**

Once the `PDS` package is installed, you will need to load the package using the `library` function.  

```{r}
library(PDS)
```

If you need directions for installing the `PDS` package, see [https://github.com/alanarnholt/PDS](https://github.com/alanarnholt/PDS). Code Books for `AddHealth`, `gapminder`, `marscrater`, and `NESARC` are in the `CodeBooks` folder of the installed `PDS` package.  To see where your system has installed the `PDS` package, type the following at the `R` prompt:

```{r eval = FALSE}
system.file(package = "PDS")
```

The Code Books are stored as pdfs.  Note: To find a word/variable in a `pdf` use `shift F` with Windows or `command F` with OSX. Type the desired word/variable inside the box and press enter/return.

Once the `PDS` pacakge is attached with the `library()` function, one may access the data in one of the data frames by typing the name of the data frame at the `R` prompt.  For example, to see the first five values in the variable `incomeperperson` from the `gapminder` data frame enter

```{r}
gapminder[1:5, 'incomeperperson']  # Show first five rows for incomeperperson
gapminder[1:5, 1:4]  # Show the first five rows for variables 1 through 4
```

### Data Management Using the `dplyr` Package

The package `dplyr` must first be installed before it can be used.  There are two ways one can install the package.  The easiest way to install `dplyr`, is to type 
```{r eval = FALSE}
install.packages("dplyr")
```
at the `R` prompt.  If you want the latest version of `dplyr`, you can install the development version from GitHub provided you have the required tools to build `R` packages installed on your machine (`Rtools` for Windows users, `XCode` for Mac users, etc.).  

```{r eval = FALSE}
if (packageVersion("devtools") < 1.6) {
  install.packages("devtools")
}
devtools::install_github("hadley/lazyeval")
devtools::install_github("hadley/dplyr")
```

### Subsetting Your Data

Create a subset of people 25 or younger who smoked in the last 12 months.  To do this, one can use the `dplyr` package. Note that `CHECK321 == 1` is used to see if subject has smoked in the past 12 months and `!is.na()` is used to make sure the subset does not contain `NA`s for the chosen variables.

```{r}
library(dplyr)
nesarc <- tbl_df(NESARC) %>% 
  filter(!is.na(CHECK321) & !is.na(AGE) & CHECK321 ==1 & AGE <= 25)
dim(nesarc)
```

The `nesarc` data frame contains `r dim(nesarc)[2]` variables.  

Three different approaches to subsetting data will be given.  The first approach is to use the `dplyr` function `filter`; the second approach is to use indices; and the third approach is to use the function `subset`.  Consider creating a subset of the `NESARC` data set where a person indicates

1. He/she has smoked over 100 cigarettes (`S3AQ1A == 1`)
2. He/she has smoked in the past year (`CHECK321 == 1`)
3. He/she has typically smoked every day over the last year (`S3AQ3B1 == 1`)
4. He/she is less than or equal to 25 years old (`AGE <= 25`)


The first approach uses the `filter` function with the `%>%` function.  Although it is not a requirement, the data frame `NESARC` is converted to a `data frame tbl` per the advice given in the `dplyr` vignette.

```{r label = "SUB1"}
NESARCsub1 <- tbl_df(NESARC) %>%
  filter(S3AQ1A == 1 & CHECK321 == 1 & S3AQ3B1 == 1 & AGE <= 25)
dim(NESARCsub1)
```

The second approach uses standard indexing.

```{r label = "SUB2"}
NESARCsub2 <- NESARC[NESARC$S3AQ1A == 1 & NESARC$CHECK321 == 1 & 
                       NESARC$S3AQ3B1 == 1 & NESARC$AGE <= 25, ]
dim(NESARCsub2)
```

The third approach uses the `subset` function.
```{r label = "SUB3"}
NESARCsub3 <- subset(NESARC, subset = S3AQ1A == 1 & CHECK321 == 1 & 
                       S3AQ3B1 == 1 & AGE <= 25)
dim(NESARCsub3)
```

Use the verb `select` from the `dplyr` package to select the variables `CHECK321`, `AGE` (age of subject), `TAB12MDX` (tobaco dependence past 12 months), `S3AQ3B1` (usual smoking quantity), `ETHRACE2A` (ethnicity), `SEX` (gender), `MAJORDEPLIFE` (Dysthymia-Lifetime) and `S3AQ3C1` (usual smoking frequency).  

### Renaming Variables

Renaming the variables `IDNUM`, `ETOTLCA2`, `ETHRACE2A`, `S3AQ3B1`, `AGE`, `MAJORDEPLIFE`, `DYSLIFE`, `TAB12MDX`, and `S3AQ3C1` to `UniqueID`, `EthanolConsumption`, `Ethnicity`, `SmokingFrequency`, `Age`, `MajorDepression`, `DysthymiaLifetime`, ` TobaccoDependence`, and `DailyCigsSmoked` respectively, is done with the `rename` function.  Although it is not a requirement, the data frame `NESARC` is converted to a `data frame tbl` per the advice given in the `dplyr` vignette.  The `rename` function takes a data frame or data frame tbl as its first argument while the syntax `new_name = old_name` is used as the second argument.

```{r label = "RENAME"}
NESARC[1:5, 1:5]  # Show first 5 rows and first 5 columns of NESARC
NESARCtbl <- tbl_df(NESARC) %>% 
  rename(UniqueID = IDNUM, EthanolConsumption = ETOTLCA2, Ethnicity = ETHRACE2A, 
         SmokingFrequency = S3AQ3B1, Age = AGE, MajorDepression = MAJORDEPLIFE, Sex = SEX, 
         DysthymiaLifetime = DYSLIFE, TobaccoDependence = TAB12MDX, DailyCigsSmoked = S3AQ3C1) %>% 
  select(UniqueID, EthanolConsumption, Ethnicity, SmokingFrequency, Age, MajorDepression, 
         DysthymiaLifetime, TobaccoDependence, DailyCigsSmoked, S3AQ1A, CHECK321, 
         SmokingFrequency, Sex)  
NESARCtbl
```

### Coding Missing Values

Note that the variable `S3AQ3B1` renamed to `SmokingFrequency` uses a 9 to record `unknown` for smoking frequency.  

```{r label = "CodeMissing"}
NESARCtbl$SmokingFrequency[NESARCtbl$SmokingFrequency == 9] <- NA
summary(NESARCtbl$SmokingFrequency)  # Note that 9 still appears
NESARCtbl$SmokingFrequency <- factor(NESARCtbl$SmokingFrequency)[, drop = TRUE]
summary(NESARCtbl$SmokingFrequency)  # Unused level no longer appears
```

I hate to look back at the code book to figure out what the heck the numbers represent in this case!  
One of the great things about the `R` language is how it handles a wide variety of data types.  
In this case, the variable `SmokingFrequency` is a factor with numeric labels for the levels of the factor.  
To change the numeric labels to text type the following:

```{r label = "Freq"}
NESARCtbl$SmokingFrequency <- factor(NESARCtbl$SmokingFrequency, 
                                     labels = c("Every Day", "5 to 6 Days/week", 
                                                "3 to 4 Days/week", "1 to 2 Days/week", 
                                                "2 to 3 Days/month", "Once a month or less"))
summary(NESARCtbl$SmokingFrequency)
xtabs(~SmokingFrequency, data = NESARCtbl) # Note how the NA's are not printed
```

The following `ggplot2` code is used to graph the values in `SmokingFrequency`.

```{r label = "ggSmoke"}
library(ggplot2)
ggplot(data = NESARCtbl, aes(x = SmokingFrequency)) + 
  geom_bar(fill = "lightgray") + 
  labs(x = "Smoking Frequency") +
  theme_bw() 
```

If you do not want to see the `NA`'s or have the labels on the $x-$ axis overlap, use

```{r label = "ggSmoke2", fig.width = 6, fig.height = 6}
ggplot(data = na.omit(NESARCtbl[ , "SmokingFrequency", drop = FALSE]), aes(x = SmokingFrequency)) + 
  geom_bar(fill = "lightgray") + 
  labs(x = "Smoking Frequency") +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 55, hjust = 1.0)) 
```

### Collapsing Categories

Consider the variable `S1Q6A` with 14 levels recording the highest level of education of the participant.  To collapse the categories in to dichotomous variable indicating the presence of a high school degree use the `ifelse` function. The levels `1`, `2`, `3`, `4`, `5`, `6`, and `7` of the variable `S1Q6A` correspond to education levels less than completing high school.

```{r label = "CollapseHS"}
NESARC$HS_DEGREE <- factor(ifelse(NESARC$S1Q6A %in% c("1", "2", "3", "4", "5", "6", "7"), 
                                  "No", "Yes"))
summary(NESARC$HS_DEGREE)
```

### Creating a Factor from a Numeric Vector

To create a factor with three levels (young adults, adults, older adults) using the variable `AGE` (a numeric vector) use the function `cut`.

```{r label = "CUT1"}
NESARC$AGEfac <- cut(NESARC$AGE, breaks = c(18, 30, 50, Inf), 
                     labels = c("Young Adult", "Adult", "Older Adult"), 
                     include.lowest = TRUE)
summary(NESARC$AGEfac)
```

Consider collapsing the numeric vector `S3AQ3C1` (usual quantity when cigarettes smoked) into a categorical variable with 5 levels.

```{r label = "CUT2"}
NESARC$S3AQ3C1fac <- cut(NESARC$S3AQ3C1, breaks = c(0, 5, 10, 15, 20, 100), 
                         include.lowest = TRUE)
summary(NESARC$S3AQ3C1fac)
```

### Aggregating Variables using `ifelse`

Consider creating a new variable `DepressLife` which is `Yes` if the variable `MAJORLIFE` is a 1 or if `DYSLIFE` is a 1.

```{r label = "Agg1"}
NESARC$DepressLife <- factor(ifelse( (NESARC$MAJORDEPLIFE == 1 | NESARC$DYSLIFE == 1), "Yes", "No"))
summary(NESARC$DepressLife)
```

As another example, consider creating a variable `PPpanic` for past and present panic disorders.  The variables `APANDX12` and `APANDXP12` record a panic disorder within the last 12 months and prior to the last 12 months with agoraphobia, respectively.  The variables `PANDX12` and `PANDXP12` record a panic disorder within the last 12 months and prior to the last 12 months without agoraphobia, respectively.

```{r label = "PANIC"}
NESARC$PPpanic <- factor(ifelse( (NESARC$APANDX12 == 1 | NESARC$APANDXP12 == 1 | 
                                    NESARC$PANDX12 == 1 | NESARC$PANDXP12 == 1 ), "Yes", "No"))
summary(NESARC$PPpanic)
```

Consider a new variable `AllDeprSymp` that records whether a subject exhibits all depression symptoms or not.  There are 19 depression symptoms recorded in the variables `S4AQ4A1` - `S4AQ4A19`.  Note that the logical operator `&` is used to test if all depression symptoms are present.

```{r label = "DEPsym"}
NESARC$AllDeprSymp <- factor(ifelse( (NESARC$S4AQ4A1 == 1 & NESARC$S4AQ4A2 == 1 & 
                                        NESARC$S4AQ4A3 == 1 & NESARC$S4AQ4A4 == 1 & 
                                        NESARC$S4AQ4A5 == 1 & NESARC$S4AQ4A6 == 1 & 
                                        NESARC$S4AQ4A7 == 1 & NESARC$S4AQ4A8 == 1 & 
                                        NESARC$S4AQ4A9 == 1 & NESARC$S4AQ4A10 == 1 & 
                                        NESARC$S4AQ4A11 == 1 & NESARC$S4AQ4A12 == 1 & 
                                        NESARC$S4AQ4A13 == 1 & NESARC$S4AQ4A14 == 1 & 
                                        NESARC$S4AQ4A15 == 1 & NESARC$S4AQ4A16 == 1 & 
                                        NESARC$S4AQ4A17 == 1 & NESARC$S4AQ4A18 == 1 & 
                                        NESARC$S4AQ4A19 == 1), "Yes", "No"))
summary(NESARC$AllDeprSymp)
```



### Creating a Composite Factor

The following code selects the 19 variables that deal with depression using the `select` function from `dplyr`. The variable `nDS` counts the number of depression symptoms a subject displays. 

```{r label = "CompositeFactor"}
mysum <- function(x){sum(x == 1)}
myadd <- function(x){apply(x, 1, mysum)}
ndf <- NESARC %>%
  select(contains("S4AQ4A"))
nDS <- myadd(ndf)
ndf <- cbind(ndf, nDS)
xtabs(~nDS, data = ndf)    
```

### Creating a New Variable with `mutate`

```{r label = "Mutate"}
MINI <- tbl_df(NESARC) %>% 
  select(S1Q24FT, S1Q24IN, S1Q24LB, SEX) %>% 
  filter(S1Q24FT < 99, S1Q24IN < 99, S1Q24LB < 999) %>% 
  mutate(Inches = (S1Q24FT*12 + S1Q24IN),
         Sex = factor(SEX, labels = c("Male", "Female"))) %>% 
  rename(Weight = S1Q24LB)
MINI
```


#### Convert a factor to numeric

Be careful when converting a factor to a vector of numeric values based on the factor labels.  One may be tempted to use `as.numeric` to convert a factor to a numeric vector.  However, using `as.numeric(my_factor)` returns a numeric vector of the index levels not the actual values.

To estimate the total number of cigarettes a subject smokes per month,  convert `S3AQ3B1` renamed to `SmokingFrequency` in `NESARCtbl` (a factor with 6 levels) to a numeric variable using `as.numeric`.  `DaysSmoke` estimates the days per month a subject smokes.  The variable `TotalCigsSmoked` estimates the monthly number of cigarettes a subject smokes per month by multiplying `DaysSmoke` times `DailyCigsSmoked` (`S3AQ3C1`).  The numeric variable `TotalCigsSmoked` is converted into a factor with roughly equivalent numbers stored in each level of the factor `CigsSmokedFac` using the `cut` function.

```{r}
summary(NESARCtbl$SmokingFrequency)
levels(NESARCtbl$SmokingFrequency)
table(as.numeric(NESARCtbl$SmokingFrequency))
```


```{r}
NESARCtbl$DaysSmoke <- as.numeric(NESARCtbl$SmokingFrequency)
NESARCtbl$DaysSmoke[NESARCtbl$DaysSmoke == 1] <- 30
NESARCtbl$DaysSmoke[NESARCtbl$DaysSmoke == 2] <- 4*5.5
NESARCtbl$DaysSmoke[NESARCtbl$DaysSmoke == 3] <- 4*3.5
NESARCtbl$DaysSmoke[NESARCtbl$DaysSmoke == 4] <- 4*1.5
NESARCtbl$DaysSmoke[NESARCtbl$DaysSmoke == 5] <- 2.5
NESARCtbl$DaysSmoke[NESARCtbl$DaysSmoke == 6] <- 1
# Using dplyr again
NESARCtbl <- NESARCtbl %>% 
  mutate(TotalCigsSmoked = DaysSmoke*DailyCigsSmoked)
proportions <- quantile(NESARCtbl$TotalCigsSmoked, na.rm = TRUE)
proportions
NESARCtbl$CigsSmokedFac <- cut(NESARCtbl$TotalCigsSmoked, breaks = proportions, 
                               include.lowest = TRUE)
NESARCtbl
```

#### Subsetting Again

```{r}
NESARCtbl <- NESARCtbl %>% 
  filter(S3AQ1A == 1 & CHECK321 == 1 & SmokingFrequency == "Every Day" & Age <= 25)
dim(NESARCtbl)
str(NESARCtbl)
summary(NESARCtbl)
```

#### Exploratory Graphs

**Numeric** data $\rightarrow$ histogram and density plots.

```{r label = "EDAquan"}
library(ggplot2)
ggplot(data = NESARCtbl, aes(x = EthanolConsumption) ) + 
  geom_histogram(binwidth = 1, fill = "pink") + 
  theme_bw() +
  labs(x = "Ethanol Consumption")
ggplot(data = NESARCtbl, aes(x = EthanolConsumption) ) + 
  geom_density(fill = "pink") + 
  theme_bw() +
  labs(x = "Ethanol Consumption")
```

**Categorical** data $\rightarrow$ bar graphs.

```{r label = "EDAqual"}
ggplot(data = NESARCtbl, aes(x = Ethnicity)) +
  geom_bar(fill = c("snow", "brown4", "red", "yellow", "tan"), color = "black") + 
  theme_bw() 
```

We should provide more descriptive labels for the factors `Ethnicity`, `MajorDepression`, `DysthymiaLifetime`, and `TobaccoDependence`.

#### Changing Levels and Labeling Variables

```{r}
NESARCtbl$TobaccoDependence <- factor(NESARCtbl$TobaccoDependence, 
                         labels = c("No Nicotine Dependence", "Nicotine Dependence"))
xtabs(~TobaccoDependence, data = NESARCtbl)
NESARCtbl$TobaccoDependence <- factor(NESARCtbl$TobaccoDependence, 
                         levels = c("Nicotine Dependence", "No Nicotine Dependence"))
xtabs(~TobaccoDependence, data = NESARCtbl)
NESARCtbl$Ethnicity <- factor(NESARCtbl$Ethnicity, 
                              labels = c("Caucasian", "African American", 
                                         "Native American", "Asian", "Hispanic"))
NESARCtbl$Sex <- factor(NESARCtbl$Sex, labels = c("Male", "Female"))
table(NESARCtbl$Sex)
NESARCtbl$Sex <- factor(NESARCtbl$Sex, levels = c("Female", "Male"))
table(NESARCtbl$Sex)
NESARCtbl$SmokingFrequency <- factor(NESARCtbl$SmokingFrequency, 
                         levels = c("Once a month or less", "2 to 3 Days/month", "1 to 2 Days/week",  
                                    "3 to 4 Days/week", "5 to 6 Days/week", "Every Day"))
NESARCtbl$MajorDepression <- factor(NESARCtbl$MajorDepression, 
                                    labels = c("No Depression", "Yes Depression"))
ggplot(data = NESARCtbl, aes(x = Ethnicity)) +
  geom_bar(fill = c("snow", "brown4", "red", "yellow", "tan"), color = "black") + 
  theme_bw() + 
  theme(axis.text.x  = element_text(angle = 55, hjust = 1)) + 
  labs(x = "")
```

```{r label = "ggSmoke2A", fig.width = 6, fig.height = 4}
T1 <- xtabs(~ TobaccoDependence + MajorDepression, data = NESARCtbl)
T1
ggplot(data = NESARCtbl, aes(x = MajorDepression, fill = TobaccoDependence)) +
  geom_bar() + 
  theme_bw()
T2 <- prop.table(T1, 2)
T2
ggplot(data = NESARCtbl, aes(x = MajorDepression, fill = TobaccoDependence)) +
  geom_bar(position = "fill") + 
  labs(x = "", y = "Fraction", 
       title = "Fraction of young adult daily smokers\nwith and without nicotine addiction\nby depression status") +
  theme_bw() +
  scale_fill_manual(values = c("red", "pink"), name = "Tobacco Addiction Status")
###
ggplot(data = NESARCtbl, aes(x = MajorDepression, fill = TobaccoDependence)) +
  geom_bar(position = "fill") + 
  labs(x = "", y = "Fraction", 
       title = "Fraction of young adult daily smokers\nwith and without nicotine addiction\nby depression status") +
  theme_bw() +
  scale_fill_manual(values = c("red", "pink"), name = "Tobacco Addiction Status") + 
  facet_grid(Sex ~ .)
```

```{r fig.width = 6.5, fig.height = 7.5}
ggplot(data = NESARCtbl, aes(x = MajorDepression, fill = TobaccoDependence)) +
  geom_bar(position = "fill") + 
  labs(x = "", y = "Fraction", 
       title = "Fraction of young adult daily smokers\nwith and without nicotine addiction\nby depression status") +
  theme_bw() +
  scale_fill_manual(values = c("red", "pink"), name = "Tobacco Addiction Status") + 
  facet_grid(Ethnicity ~ Sex) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### kable

```{r, echo = FALSE, results = 'asis'}
knitr::kable(NESARCtbl[1300:1305, c("EthanolConsumption", "Sex", "MajorDepression")], 
             align = c("c","c","c"), caption = "Three Selected Columns")
```

```{r, echo = FALSE, results = 'asis'}
knitr::kable(NESARCtbl[1300:1305, c("EthanolConsumption", "Sex", "MajorDepression")], 
             align = c("c","c","c"), caption = "Three Selected Columns", col.names = c("Ethanol Consuption", "Gender", "Depression"))
```

#### pander

```{r, echo=FALSE, results='asis'}
pander::pandoc.table(NESARCtbl[1300:1305, c("EthanolConsumption", "Sex", "MajorDepression")],
                     caption = "Three Selected Columns")
```


